From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-26 upi.tex
reorganizing

The proceedings of the 2012 Symposium on Principles of Programming
Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
\emph{Information Effects} by James and Sabry and \emph{Canonicity for
  2-dimensional type theory} by Licata and Harper. The first paper, motivated by
the physical nature of
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
proposed a reversible language $\Pi$ in which every program is a type
isomorphism. The second paper, motivated by the topological nature of
computation~\cite{vv06,hottbook}, proposed a judgmental formulation of
intensional dependent type theory with a twice-iterated identity type. During
the conference, it became apparent, at an intuitive and informal level, that the
two papers had strong similarities. Formalizing the precise connection was far
from obvious, however. In this paper we report on a formal connection between
appropriately formulated reversible languages on one hand and univalent
universes on the other.

@@

The practice of programming languages is replete with \emph{ad hoc} instances of
reversible computations: database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility. But intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
and by the potential for efficient quantum
computation~\cite{springerlink:10.1007/BF02650179}.

%%%%%
\subsection{The $\Pi$ Family of Languages}

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. While at first sight,
$\Pi$ might appear \emph{ad hoc},~\cite{Carette2016} shows that it arises naturally
from an ``extended'' view of the Curry-Howard Correspondance: rather than looking
at mere \emph{inhabitation} as the main source of analogy between logic
and computation, \emph{type equivalences} becomes the source of analogy.  This
allows one to see an analogy between algebra and reversible computation.
Furthermore, this works at multiple levels: that of $1$-algebra (types form a
semiring under isomorphism), but also $2$-algebra (types and equivalences form a
weak Rig Groupoid).  In other words, by taking ``weak Rig Groupoid'' as the
starting semantics, one naturally gets $\Pi$ as the syntax for the language of
proofs of isomorphisms -- in the same way that many terms of the
$\lambda$-calculus arise from Cartesian Closed Categories.

We will later identify a ``featherweight'' version of $\Pi$ to use in our formal
development.  But first we illustrate the flavor of the family of languages with
some examples.

@@

\begin{comment}
In this paper we will
identify a ``featherweight'' representative to use for the formal
development. But first, we illustrate the flavor of the family of languages with
some examples written in a Haskell syntax. First consider the following program:

{\footnotesize
\begin{verbatim}
addSub1 :: (Int, Int) :<=> (Int, Int)
addSub1 = CommuteTimes
      :.: (UnfoldN :*: Id)
      :.: Distribute
      :.: (Id :+: CommuteTimes)
      :.: Factor
      :.: (FoldN :*: Id)
\end{verbatim}}
%
\noindent which takes a pair of natural numbers and processes them as follows depending on
whether the second number is 0 or not. If the second number is 0, the evaluation
proceeds as follows:
$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
applications of this program would increment the first number and decrement the
second number maintaining the sum at all times. When the second number reaches
0, the numbers are swapped.

As another example, consider the following expressions culminating with the
implementation of the Toffoli gate, which is universal for combinational
circuits. We have two implementations of simple boolean negation: a direct one
and a more convoluted one. These two implementations are ``equivalent'' in the
sense to be explained below. The expression \verb|cond| is the main reversible
control flow mechanism: it allows for the conditional execution of one of two
branches but retains the choice of which branch was executed. This mechanism can
be used twice to implement the Toffoli gate, which performs a conditional
negation on one bit if two other bits are both true:

{\footnotesize
\begin{verbatim}
inot :: Bool :<=> Bool
inot = CommutePlus

inot2 :: Bool :<=> Bool
inot2 = UnitTimesI
    :.: CommuteTimes
    :.: (CommutePlus :*: Id)
    :.: CommuteTimes
    :.: UnitTimesE

cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
cond f g = Distribute
       :.: ((Id :*: f) :+: (Id :*: g))
       :.: Factor

controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
controlled f = cond f Id

cnot :: (Bool, Bool) :<=> (Bool, Bool)
cnot = controlled inot

toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
\end{verbatim}}

Although writing circuits using the raw syntax for combinators above is tedious,
the examples illustrate the programming language nature of this family of $\Pi$
languages. In other work, one can find a compiler from a conventional functional
language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
Haskell-like surface language~\cite{theseus} which can ease writing
programs. All that reinforces the claim that this is a practical approach to
reversible programming.

The raw syntax for combinators has, however, a significant benefit: it allows
algebraic manipulation of the programs using a second level of combinators. As
an example it is possible to show the the two implementations of boolean
negation are equivalent as follows:

{\footnotesize
\begin{verbatim}
negOpt :: inot2 :<===> inot
negOpt = (Id :>: AssocLeftEquiv)
    :..: (Id :>: (SlideCommuteTimes :>: Id))
    :..: (Id :>: AssocRightEquiv)
    :..: (Id :>: (Id :>: AssocLeftEquiv))
    :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
    :..: (Id :>: (Id :>: IdElimL))
    :..: AssocLeftEquiv
    :..: (SlideUnitI :>: Id)
    :..: AssocRightEquiv
    :..: (Id :>: InvInvEquivId)
    :..: IdElimR
\end{verbatim}}
%
\noindent This program, which can be thought of as an optimization or a proof,
relates the two implementations of boolean negation via a series of elementary
optimizations of which the following are the most interesting:

{\footnotesize
\begin{verbatim}
AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
IdElimL           ::                     (Id :.: c) :<===> c
SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
\end{verbatim}}

\end{comment}
%%%%%
\subsection{$\Pi_{\mathbb{2}}$}

We will however show, in a very clear sense, how a $\Pi$-based language is
``equivalent'' to a univalent HoTT universe that possesses all the required
properties of paths. The essence of the construction can be demonstrated with a
tiny ``featherweight'' variant of $\Pi$ that includes just one type: booleans,
and a tiny ``featherweight'' univalent universe to be formalized in the next
section. Here we give a formalization of the small $\Pi$-based language which we
will use in the formalization in the rest of the paper.

The restriction of $\Pi$ to the case of just booleans is the following:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Univalent Type Families}

@@

\begin{verbatim}
Thm : Let P : Type -> Type such that for all X : Type, P(X) is a proposition. Then Σ(X,P) is the base of a univalent fibration.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Correspondence}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

>>>>

The proceedings of the 2012 Symposium on Principles of Programming
Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
\emph{Information Effects} by James and Sabry and \emph{Canonicity for
  2-dimensional type theory} by Licata and Harper. The first paper, motivated by
the physical nature of
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
proposed, among other results, a reversible language $\Pi$ in which every
program is a type isomorphism. The second paper, motivated by the connections
between homotopy theory and type theory~\cite{vv06,hottbook}, proposed a
judgmental formulation of intensional dependent type theory with a
twice-iterated identity type. During the presentations and ensuing discussions
at the conference, it became apparent, at an intuitive and informal level, that
the two papers had strong similarities. Formalizing the precise connection was
far from obvious, however.

In this paper we report on a formal connection between appropriately formulated
reversible languages on one hand and univalent universes on the other. In the
next section, we give a rational reconstruction of $\Pi$ focusing on a small
``featherweight'' fragment. In Sec.~\ref{sec:univalent}, we review
\emph{univalent fibrations} which allow us to give formal presentations of
``small'' univalent universes. In Sec.~\ref{sec:connection} we state and prove
the formal connection between the systems presented in the preceding two
sections. Sec.~\ref{sec:conclusion} puts our work in a larger context, discusses
related and future work, and concludes.

@@

Starting from the physical principle of ``conservation of
information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative}, James and
Sabry~\cite{James:2012:IE:2103656.2103667} propose a family of programming
languages $\Pi$ in which computation preserve information. Technically,
computations are \emph{type isomorphisms} which, at least in the case of finite
types, clearly preserve entropy in the information-theoretic sense.

We will identify a ``featherweight'' version of $\Pi$ to use in our formal
development.  But first we illustrate the flavor of the family of languages with
some examples.

@@

% \begin{comment}
% In this paper we will
% identify a ``featherweight'' representative to use for the formal
% development. But first, we illustrate the flavor of the family of languages with
% some examples written in a Haskell syntax. First consider the following program:

% {\footnotesize
% \begin{verbatim}
% addSub1 :: (Int, Int) :<=> (Int, Int)
% addSub1 = CommuteTimes
%       :.: (UnfoldN :*: Id)
%       :.: Distribute
%       :.: (Id :+: CommuteTimes)
%       :.: Factor
%       :.: (FoldN :*: Id)
% \end{verbatim}}
% %
% \noindent which takes a pair of natural numbers and processes them as follows depending on
% whether the second number is 0 or not. If the second number is 0, the evaluation
% proceeds as follows:
% $(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
% \mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
% second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
% $(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
% \mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
% (\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
% applications of this program would increment the first number and decrement the
% second number maintaining the sum at all times. When the second number reaches
% 0, the numbers are swapped.

% As another example, consider the following expressions culminating with the
% implementation of the Toffoli gate, which is universal for combinational
% circuits. We have two implementations of simple boolean negation: a direct one
% and a more convoluted one. These two implementations are ``equivalent'' in the
% sense to be explained below. The expression \verb|cond| is the main reversible
% control flow mechanism: it allows for the conditional execution of one of two
% branches but retains the choice of which branch was executed. This mechanism can
% be used twice to implement the Toffoli gate, which performs a conditional
% negation on one bit if two other bits are both true:

% {\footnotesize
% \begin{verbatim}
% inot :: Bool :<=> Bool
% inot = CommutePlus

% inot2 :: Bool :<=> Bool
% inot2 = UnitTimesI
%     :.: CommuteTimes
%     :.: (CommutePlus :*: Id)
%     :.: CommuteTimes
%     :.: UnitTimesE

% cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
% cond f g = Distribute
%        :.: ((Id :*: f) :+: (Id :*: g))
%        :.: Factor

% controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
% controlled f = cond f Id

% cnot :: (Bool, Bool) :<=> (Bool, Bool)
% cnot = controlled inot

% toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
% toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
% \end{verbatim}}

% Although writing circuits using the raw syntax for combinators above is tedious,
% the examples illustrate the programming language nature of this family of $\Pi$
% languages. In other work, one can find a compiler from a conventional functional
% language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
% technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
% Haskell-like surface language~\cite{theseus} which can ease writing
% programs. All that reinforces the claim that this is a practical approach to
% reversible programming.

% The raw syntax for combinators has, however, a significant benefit: it allows
% algebraic manipulation of the programs using a second level of combinators. As
% an example it is possible to show the the two implementations of boolean
% negation are equivalent as follows:

% {\footnotesize
% \begin{verbatim}
% negOpt :: inot2 :<===> inot
% negOpt = (Id :>: AssocLeftEquiv)
%     :..: (Id :>: (SlideCommuteTimes :>: Id))
%     :..: (Id :>: AssocRightEquiv)
%     :..: (Id :>: (Id :>: AssocLeftEquiv))
%     :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
%     :..: (Id :>: (Id :>: IdElimL))
%     :..: AssocLeftEquiv
%     :..: (SlideUnitI :>: Id)
%     :..: AssocRightEquiv
%     :..: (Id :>: InvInvEquivId)
%     :..: IdElimR
% \end{verbatim}}
% %
% \noindent This program, which can be thought of as an optimization or a proof,
% relates the two implementations of boolean negation via a series of elementary
% optimizations of which the following are the most interesting:

% {\footnotesize
% \begin{verbatim}
% AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
% SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
% IdElimL           ::                     (Id :.: c) :<===> c
% SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
% \end{verbatim}}

% \end{comment}

%%%%%
\subsection{$\Pi_{\mathbb{2}}$}

We present a small $\Pi$-based language which we will use in the formalization
in the rest of the paper. The restriction of $\Pi$ to the case of just booleans
is the following:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Univalent Fibrations}
\label{sec:univalent}

@@

\begin{verbatim}
Thm: Let P : Type -> Type such that for all X : Type, P(X)
is a proposition. Then Σ(X,P) is the base of a univalent
fibration.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Correspondence}
\label{sec:connection}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion, Related Work, and Conclusion}
\label{sec:conclusion}

\paragraph*{Reversible Languages.}
\noindent The practice of programming languages is replete with \emph{ad hoc}
instances of reversible computations: database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility. But intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
and by the potential for efficient quantum
computation~\cite{springerlink:10.1007/BF02650179}.

\paragraph*{The $\Pi$ Family of Languages}
\noindent In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. While at first sight,
$\Pi$ might appear \emph{ad hoc},~\cite{Carette2016} shows that it arises
naturally from an ``extended'' view of the Curry-Howard correspondance: rather
than looking at mere \emph{inhabitation} as the main source of analogy between
logic and computation, \emph{type equivalences} becomes the source of analogy.
This allows one to see an analogy between algebra and reversible computation.
Furthermore, this works at multiple levels: that of $1$-algebra (types form a
semiring under isomorphism), but also $2$-algebra (types and equivalences form a
weak Rig Groupoid).  In other words, by taking ``weak Rig Groupoid'' as the
starting semantics, one naturally gets $\Pi$ as the syntax for the language of
proofs of isomorphisms -- in the same way that many terms of the
$\lambda$-calculus arise from Cartesian Closed Categories.
